Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Guideline for International Students

### 2024-01-27 demystify.tex
updates

\renewcommand{\figurename}{Fig.}
\renewcommand{\tablename}{Tab.}
\def\Section{\S}
\newcommand{\mycomment}[3][\color{blue}]{{#1{{#2}: {#3}}}}
\newcommand{\tvn}[1]{\mycomment{TVN}{#1}}{}
\newcommand{\didi}[1]{\mycomment{Didier}{#1}}{}
\newcommand{\tl}[1]{\mycomment{ThanhLe}{#1}}{}
\newcommand{\red}[1]{{\color{red}{#1}}}
\newcommand{\xz}[1]{\mycomment{Xiaokuan}{[#1]}}{}

@@

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the PhD admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While this document is primarily intended for students interested in CS, it may be relevant to students from various disciplines. 
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) universities are classified as R1.} institutions in the US (and universities in other countries).

@@

The main points of this guideline are summarized here. This gives you an overview to decide which specific topics you want to explore more thoroughly.

\begin{enumerate}
  \item Should you apply?
        \begin{itemize}
          \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than non-US schools (\S\ref{sec:should}).
        \end{itemize}
  \item How is your application evaluated?
        \begin{itemize}
          \item Applications are evaluated by the \emph{PhD Admission} committee and each application is reviewed by typically three faculty (\S\ref{sec:evalapps}).
          \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so contact and introduce yourself (\S\ref{sec:contact}).
        \end{itemize}
  \item Application Materials
        \begin{itemize}
          \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, personal statement, and recommendation letters.
          \item LORs are very important, but only if they are personalized and talk about your research ability (\S\ref{sec:lor}).
          \item Personal statement is very important. Write it in such a way that makes you \emph{stand out} (\S\ref{sec:research-statement} and \S\ref{sec:improve-your-chance})
          \item GRE \emph{is not} required (\S\ref{sec:grades}). Spend your time on something else.
          \item Grades are important, but depend on the reputation of your school (\S\ref{sec:grades}).
          \item Getting an interview is typically a \emph{good sign} (no one wants to interview weak candidates) (\S\ref{sec:interview}).
        \end{itemize}
  \item What to do after getting admitted?
        \begin{itemize}
                \item  Celebrating! Now it is your turn to evaluate the school!
          \item Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now (\S\ref{sec:accepted}).
        \end{itemize}
  \item Funding
        \begin{itemize}
          \item TA and RA are two main funding sources.  TA (teaching assistantship) is provided by the department to help profs. with classes (e.g., grading). RA (research assistantship) is provided by profs. to help them with their research (\S\ref{sec:funding}).
        \end{itemize}
  \item Choosing School and Professors
        \begin{itemize}
          \item Many schools do not offer PhD studies in CS and many CS professors \emph{cannot} advise or graduate PhD students  (\S\ref{sec:schoolsandprofs}).
          \item Contacting a prof. for research opportunities is recommended, but do it \emph{properly} (\S\ref{sec:contact}).
        \end{itemize}
  \item Miscs and FAQS
        \begin{itemize}
          \item Increasing your admission chance by being unique and standing out (\S\ref{sec:improve-your-chance}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\S\ref{sec:non-stem}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\S\ref{sec:non-us-differences}).
          \item Your 5--6 year PhD program costs about \$400K in total, but you \emph{do not} pay for it (\S\ref{sec:ra-cost}).
          \item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\S\ref{sec:happy}).
        \end{itemize}
\end{enumerate}

@@

First, I want to emphasize that PhD students in Computer Science \emph{do not} need to worry about funding, especially at good R1
universities in the US. If you are admitted, you will almost certainly
receive \emph{full funding} to support your study, including tuition,
health insurance, and stipend (monthly salary). Moreover, depending on the university,
you may even receive additional benefits such as summer pay, laptops, (conference/workshop) traveling. \S\ref{sec:funding} provides more details on funding.

Second, applying to a good US university \emph{should not} be any harder than at schools in other countries. In fact, it might even be more flexible since CS PhD in the US \emph{do not} require having an MS or a topic/proposal/adviser in advance. If you believe you have a chance in other countries, e.g., South Korea, Singapore, Germany, UK, Japan, and Australia, then you will surely have a chance in the US as well. \S\ref{sec:non-us-differences} compares CS PhD study in the US to other countries.

\begin{commentbox}[Vu:]
  One of the reasons I create this document is that my colleagues at GMU are interested in recruiting Vietnamese students and are surprised when seeing very few applications in Vietnam (see \S\ref{sec:ack}). Each year our CS program receives more than 350 PhD applications, most of which are international but only 3--4 are from Vietnam. In general the number of CS PhD applications from Vietnam to US universities is very low and more would be welcome.
\end{commentbox}

\begin{domesticbox}[Domestic students:] If you're a domestic student, you have several advantages in your application.  First, we already know about your school (see \S\ref{sec:your-school}), and you are also more familiar with the US education system and culture.  You also have more opportunities for funding, e.g., through government scholarships for US citizens and residents. Finally, compared to international students, PhD applications in CS from domestic students are few and that can really help your case.
\end{domesticbox}

@@

The PhD adcom typically involves assistant professors in the department (see \S\ref{sec:faculty-types} for various faculty titles). This provides junior faculty the opportunities to recruit students. The chair of the committee will be a senior professor, but they likely will not review individual applications and instead assign them to committee members. The chair will look at various factors such as research interests or mentioning faculty names to assign the applications to appropriate faculty.

At GMU, we usually decide that a full-time PhD candidate is either (i) admit with funding (TA or RA, see \S\ref{sec:funding}) or (ii) rejected. In other words, in most cases, we either
admit you with full funding, or we reject your application. In some rare cases, we admit
without funding because you have funding on your own, e.g.,
supported by your government or having external fellowships. We justify
our decision with a summary of your application, where we list
strengths (e.g., a well-known school) and weaknesses (e.g., weak
LORs).

@@

However, it is understandable that many students do not have the opportunities to publish in top places. Thus, general writings, even those under submissions or rejected, would still be good (and much better than nothing).  But be sure to upload your papers with your application and talk about them in your statement (see \S\ref{sec:research-statement}).  Note that local conferences and non-English journals or conferences do
not carry as much weight since their quality is often unknown to US faculty. However, if you have published in such places, you should still mention them in your statement and explain why they are good.

% In CS, conferences are important and top conferences are prestigious. You can find top CS conferences at places such as \href{https://csrankings.org}{CSRankings}, which ranks CS programs based on how their faculty publish at top \emph{conferences} (see more in \S\ref{sec:ranking}).

@@

  Sometimes students would go through great length just to get letters from well-known professors in their school (e.g., department head). But as mentioned, if these professors do not know you, the letters are generic and carry little value, in fact, \red{red flags}. Moreover, a top professor in Vietnam might not be well-known to US faculty (see more details in \S\ref{sec:your-school}). So save the trouble and just get letters from \emph{any} professors/supervisors who knows you well and can write a good letter about \emph{your} research ability. It's better to have a good personalized
  letter about your own research ability from someone who is less
  well-known than a generic/weak letter from a well-known person.

@@

While you might not have control over LORs or \hyperref[sec:your-school]{where your go to school}, you do over your
statement of purpose (SOP) or personal statement\footnote{Some schools separate these documents and ask you to write both: SOP, which focuses on research experiences, and Personal Statement, which is everything more personal, e.g., why PhD, challenges, diversity, etc}! So write it well because it could make a big difference.
In your statement, you have the opportunity to make your application stand out and unique, which can make you \emph{fit} the CS program you're applying to, even if you don't have very strong research experience.
A well-written SOP also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \S\ref{sec:funding}).
% is important because if you need (GTA) funding, it will provide evidence
% that you can teach and communicate with students.

There are various guides on writing statement, e.g.,~\cite{blattman2022writing}, and \href{https://cs-sop.org/}{many example statements} are available. So I will not talk too much about statements. In short, focus on your research goal and vision, and convince us that you can achieve it through your experience, e.g., published papers, or if you work on some projects by yourself, talk about it. Also, talk about things that adcom members might not know about, e.g., your Github project with 1K+ stars or your regular contributions to well-known open-source projects. Finally, talk about what makes you different than the other 1000 applicants that we are considering (e.g., see \S\ref{sec:improve-your-chance} for increasing your admission chance).

@@

\begin{commentbox}[Vu:]
  I often read the research statement (and the LORs) first. If I am
  persuaded by then, I would skim over other factors and advocate for
  admission (unless I see red flags in other parts). If I am not
  convinced, then I will likely recommend rejection (unless I see
  something standout in other parts).
  \tcblower
  Do careful research on professors, don't mention \emph{emeritus} or  adjunct faculty (see more about various types of faculty in \S\ref{sec:faculty-types}).
  Also, be careful not to send statements to wrong schools or mixing
  facts (e.g., talking about school X but mentioned about working with
  profs. at school Y; and definitely do not talk about George Washington when applying to George Mason). I have seen such statements more time that I
  should.
\end{commentbox}

@@

Graduating from top universities \emph{that we recognize} helps. For example, if your school is well-known, then it is \emph{``top foreign''}, which is definitely a plus.
However, if committee members do not know much about schools in your country, they will likely treat your school as
\emph{``unknown foreign''}, which can be a minus point because they are uncertain about the quality of your school.

@@

This can be an issue for students in many top international universities where the competition is so high that very good students can still have low rankings from these schools (and be overlooked by the adcom).
So what to do with this? as mentioned in \S\ref{sec:your-school} you can put a note about this in your statement and ask your LoR writers to talk about it.

@@

In short, getting an interview is a good sign; it means that someone is considering you. If they are not interested in your application, they will not proceed with an interview.
However, do not worry much if you do not get an interview! For example, at GMU, most of our admitted students with TA do not go through any interviews.

@@

        You can then share this sheet with your reference writers and remind them to submit LoRs on your behalf (see \S\ref{sec:lor}).
\end{itemize}

@@

By around March you should hear back from most PhD programs you applied to. If you haven't heard back, reach out via email and ask about the status of your application.
If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!  Important factors to consider are the reputation of schools and professors (\S\ref{sec:schoolsandprofs}), and funding availability (\S\ref{sec:funding}). You will have to make your decision by around \emph{April 15}.

@@

As mentioned, if you're admitted to a \emph{good} CS PhD program, you should not have to worry about funding!  
In the US, the common types of funding for PhD are \emph{graduate teaching assistant} (GTA or TA), \emph{graduate research assistant} (GRA or RA), and \emph{Fellowship}.
RA is paid by a prof. for you to do their research. TA is paid by the department for you to help with teaching. Finally, fellowship is an independent funding that can come from a school, a company, or an organization. Tab.~\ref{tab:funding} summarizes the differences.
Note that funding is typically more available for PhD students than 
Masters. %Undergraduates typically have no funding and also have to pay international tuition, which is very expensive!.

@@

For a full breakdown of how much a graduate student costs, see \S\ref{sec:ra-cost}.

@@

\subsection{Research Assistant (RA)}
RA is provided through a professor through their own funding so you can work on their project.  
You do not need to teach as an RA, so you can focus on your research. Depending on the professor, RA may be available during the summer. \S\ref{sec:ra-cost} gives more details on RA budget.

\textbf{How to get RA?} When a professor recruits you, they will likely give you RA right away (e.g., when you apply).  A common scenario is that you first get admitted with TA, and then after a year or two find an adviser to support you with RA. See how to contact a prof. in \S\ref{sec:contact} for research opportunities.

@@

Thus, if your goal is Ph.D. in CS, you have to aim only for schools offering such a degree.  %This might sound obvious but many students (even those in the US) get confused due to the large number of universities. 
While this can be confusing due to the large number of universities in the US, a little research, e.g., searching for PhD in CS from the school website, will help you find out. All schools listed in \S\ref{sec:ranking} have PhD studies in CS, so you can start there.

@@

You can learn about CS programs and research expertise of faculty using resources such as \href{https://csrankings.org}{CSRankings.org}, which is designed specifically to help prospective PhD students in Computer Science!  You will be very surprised to learn that a school that you didn't know much about can have very strong research in your interested topic (and vice versa, a school you thought highly about might have no faculty working in the research field you're interested in). This is also a good way to learn about individual faculty (who works on what) and well-known CS conferences\footnote{In CS (and probably only in CS), conferences, not journals, are often the main venue to publish research finding.}. \S\ref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.

@@

However, in general, rankings can be superficial and you need to do more research to be informed and make better decision. For example, if you get admissions to several places, you should consider attending Open Houses and contact profs. that you're interested in at those place and talk to them. They would be more willing to chat with you now that you have been admitted.  Ask them questions about their work, how they manage students, their expectations. You can even ask to contact their students. See more in \S\ref{sec:accepted} on what to do after getting admitted.

@@

\section{Choosing an Adviser}
Obviously, there is no one-size-fit-all answer to this question. The best adviser is the one that you can work well with, i.e., \emph{fits} you.  This is difficult and takes time. Fortunately, unlike many countries that require finding an adviser and research topic before starting the PhD (\S\ref{sec:non-us-differences}), CS PhD programs in the US will typically give you a couple of years to search for advisers and research topics.  This is especially true if you're admitted with TA (\S\ref{sec:ta}), which gives you time to explore and find adviser.

@@

Here are some general advice to find professor.  First, search for profs. that share similar research interests. You can learn about faculty and their level of research activity through the faculty's website or even from CSRankings.  For example, in CSRankings, if you want to work with PL, you can search for those publish in PL conferences.  If you want to work with SE \emph{and} AI, you can search for faculty who work in both SE and AI.   You can also go to their website, look at their papers and projects.  In any case, you should contact the professor and talk to them. For example, you can (cold) email them and ask about their research and if they are taking new students (\S\ref{sec:contact}), and schedule appointments to chat with them (office hrs are often the best time).

@@

  In other fields, e.g. Software Engineering, Vietnamese students face challenges in reaching US professors. Do you have any tips for Vietnamese students who want to connect with US professors and work as research assistants?
  \tcblower
  \textbf{Vu}: \S\ref{sec:contact} shows how to contact a professor for research opportunities. Many probably will say no (or do not reply) as they do not have the bandwidth to take on random students, but some may say yes if they see potential fit.
\end{commentbox}

@@

Not every faculty can serve as your formal adviser. Let's try to understand the different types of faculty and their roles. For example, you might hear about tenured,  tenure track, teaching faculty.  You might also things such as assistant, associate, full, adjunct, emeritus, teaching, research professors. Here is a primer on these terminologies.

\paragraph{Tenure-line and teaching faculty} Tenure line faculty, consisting of tenured and tenured-track profs., focus on research, which includes publishing papers, obtaining grants, mentoring Ph.D students.  They often have very low teaching load (e.g., 1 per semester). \emph{Tenured} faculty are professors who have been promoted to have a permanent position (informally, very hard to fire them).   \emph{Tenure-track} faculty are (more likely young) professors who are on the track to get tenure.  \textbf{Both tenure and tenured track faculty can serve as a formal adviser of Ph.D. students}. This is very important because you need to have a tenure line faculty as your main PhD adviser. \S\ref{sec:tenure-vs-tenure-track} talks more about choosing between tenure and tenured-track professor as your adviser.

In contrast, \emph{teaching} (or instructional faculty or professor of practice) mainly focus on teaching. They typically teach 3--4 classes per semester (which is quite a lot) and \emph{do not} have research responsibilities (e.g., they do not have to worry much about publishing papers, obtaining grants).  They can mentor Ph.D. students but they typically cannot serve as a formal adviser of Ph.D. students (i.e., they \emph{cannot graduate} PhD students).

It is worth noting that professors in a non-CS department also unlikely can serve as formal PhD advisers for CS students (see \S\ref{sec:related-fields}).

\paragraph{Titles} Faculty have titles, e.g., assistant, associate, full, regardless if they are tenure line or teaching.  Assistant means new faculty, associate means they have been promoted, and full means they are senior. A tenure track faculty starts with being an assistant professor and then gets tenure and promoted to associate (typically after 6 years) and then full (time varies, some become full within 3-4 years, some 10+ years, some remain associate).

Adjunct faculty not full-time faculty, e.g., they might be working in industry and teach a class or two for fun.  Emeritus means they are retired but still have some affiliation with the university.  Research faculty (or scientist) are typically non-tenure line faculty who focus on research.  Due to their roles, adjunct, emeritus, and research faculty typically do not advise Ph.D. students.

@@

First, if you want to contact a prof. to \emph{ask about your admission chance}, please \textbf{don't}. We don't know and can't answer because as explained in \S\ref{sec:evalapps}, we don't make individual decisions and might not even be assigned to evaluate your application.  It is the same as sending a paper draft to a journal editor and ask them if your paper has a chance.

>>>>

\renewcommand{\figurename}{Fig.}
\renewcommand{\tablename}{Tab.}
\def\Section{\S}
\renewcommand{\figureautorefname}{Fig.}
\renewcommand{\tableautorefname}{Tab.}
\makeatletter
\renewcommand{\chapterautorefname}{\S\@gobble}
\renewcommand{\sectionautorefname}{\S\@gobble}
\renewcommand{\subsectionautorefname}{\S\@gobble}
\makeatother

\newcommand{\mycomment}[3][\color{blue}]{{#1{{#2}: {#3}}}}
\newcommand{\tvn}[1]{\mycomment{TVN}{#1}}{}
\newcommand{\didi}[1]{\mycomment{Didier}{#1}}{}
\newcommand{\tl}[1]{\mycomment{ThanhLe}{#1}}{}
\newcommand{\red}[1]{{\color{red}{#1}}}
\newcommand{\xz}[1]{\mycomment{Xiaokuan}{[#1]}}{}

@@

% though \emph{very} top schools could be very selective, e.g., see the \href{https://da-data.blogspot.com/2015/03/reflecting-on-cs-graduate-admissions.html}{admission process} at CMU
So I want to share some details about the admission process and advice for those who are interested in applying for a \textbf{PhD in Computer Science in the US}.
While this document is primarily intended for students interested in CS, it might be relevant to students from various disciplines.
Furthermore, although many examples are specifics for schools that I and other contributors of this document know about, the information should be generalizable to other R1\footnote{An \href{https://en.wikipedia.org/wiki/List_of_research_universities_in_the_United_States}{R1 institution} in the US is a research-intensive university with a high level of research activity across various disciplines. Currently, 146 (out of 4000) universities are classified as R1.} institutions in the US (and universities in other countries).

@@

Below we summarize the main points of this guideline. This gives you an overview to decide which specific topics you want to explore more thoroughly.

\begin{enumerate}
  \item Should you apply?
        \begin{itemize}
          \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than non-US schools (\autoref{sec:should}).
        \end{itemize}
  \item How is your application evaluated?
        \begin{itemize}
          \item Applications are evaluated by the \emph{PhD Admission} committee and each application is reviewed by typically three faculty (\autoref{sec:evalapps}).
          \item Individually faculty \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so do contact and introduce yourself (\autoref{sec:contact}).
         \end{itemize}
  \item Application Materials
        \begin{itemize}
          \item The committee will look at various factors, but the most important ones are research ability, e.g., publications, statement of purpose (SOP), and  letters of recommendation (LORs).
          \item LORs are very important, but only if they are personalized and talk about your research ability (\autoref{sec:lor}).
          \item SOP is very important. Write it in such a way that makes you \emph{stand out} (\autoref{sec:research-statement} and \autoref{sec:improve-your-chance})
          \item GRE \emph{is not} required (\autoref{sec:grades}). Spend your time on something else!
          \item Grades are important, but depend on the reputation of your school (\autoref{sec:grades}).
          \item Getting an interview is typically a \emph{good sign} as no one wants to interview weak candidates (\autoref{sec:interview}).
        \end{itemize}
  \item What to do after getting admitted?
        \begin{itemize}
                \item  Celebrating! Now it is your turn to evaluate the school!
          \item Attend \emph{Open House} to learn more about the place and \emph{interview} profs---they would be much more willing to talk to you now (\autoref{sec:accepted}).
        \end{itemize}
  \item Funding
        \begin{itemize}
          \item TA and RA are two main funding sources.  TA (teaching assistantship) is provided by the department to help profs. with classes (e.g., grading). RA (research assistantship) is provided by profs. to help with their research (\autoref{sec:funding}).
        \end{itemize}
  \item Choosing School and Professors
        \begin{itemize}
          \item Many schools do not offer PhD studies in CS and many CS professors \emph{cannot} formally advise or graduate PhD students  (\autoref{sec:schoolsandprofs}).
          \item Contacting a prof. for research opportunities is recommended, but do it \emph{properly} (\autoref{sec:contact}).
        \end{itemize}
  \item Miscs and FAQS
        \begin{itemize}
          \item Increasing your admission chance by being unique and standing out (\autoref{sec:improve-your-chance}).
          \item You can successfully apply to CS PhD even if you have non-STEM background (\autoref{sec:non-stem}).
          \item Compared to other countries, CS PhD in the US does not require an MS degree but has longer PhD study time (\autoref{sec:non-us-differences}).
          \item Your 5--6 year PhD program costs about \$400K in total, but you \emph{do not} pay for it (\autoref{sec:ra-cost}).
          \item Despite some miserable stories on social media, many PhD students have good mentors, supportive lab mates, healthy working environment ... and are happy (\autoref{sec:happy}).
        \end{itemize}
\end{enumerate}

@@

First, I want to emphasize that PhD students in Computer Science \emph{do not} need to worry about funding, especially at good R1
universities in the US. If you are admitted, you will almost certainly
receive \emph{full funding} to support your study, including tuition,
health insurance, and stipend (monthly salary). Moreover, depending on the university,
you may even receive additional benefits such as summer pay, laptops, (conference/workshop) traveling. \autoref{sec:funding} provides more details on funding.

Second, applying to a good US university \emph{should not} be any harder than at schools in other countries. In fact, it might even be more flexible since CS PhD in the US \emph{do not} require having an MS or a topic/proposal/adviser in advance. If you believe you have a chance in other countries, e.g., South Korea, Singapore, Germany, UK, Japan, and Australia, then you will surely have a chance in the US as well. \autoref{sec:non-us-differences} compares CS PhD study in the US to other countries.

\begin{commentbox}[Vu:]
  One of the reasons I create this document is that my colleagues at GMU are interested in recruiting Vietnamese students and are surprised when seeing very few applications in Vietnam (see \autoref{sec:ack}). Each year our CS program receives more than 350 PhD applications, most of which are international but only 3--4 are from Vietnam. In general the number of CS PhD applications from Vietnam to US universities is very low and more would be welcome.
\end{commentbox}

\begin{domesticbox}[Domestic students:] If you're a domestic student, you have several advantages in your application.  First, we already know about your school (see \autoref{sec:your-school}), and you are also more familiar with the US education system and culture.  You also have more opportunities for funding, e.g., through government scholarships for US citizens and residents. Finally, compared to international students, PhD applications in CS from domestic students are few and that can really help your case.
\end{domesticbox}

@@

The PhD adcom typically involves assistant professors in the department (see \autoref{sec:faculty-types} for various faculty titles). This provides junior faculty the opportunities to recruit students. The chair of the committee will be a senior professor, but they likely will not review individual applications and instead assign them to committee members. The chair will look at various factors such as research interests or mentioning faculty names to assign the applications to appropriate faculty.

At GMU, we usually decide that a full-time PhD candidate is either (i) admit with funding (TA or RA, see \autoref{sec:funding}) or (ii) rejected. In other words, in most cases, we either
admit you with full funding, or we reject your application. In some rare cases, we admit
without funding because you have funding on your own, e.g.,
supported by your government or having external fellowships. We justify
our decision with a summary of your application, where we list
strengths (e.g., a well-known school) and weaknesses (e.g., weak
LORs).

@@

However, it is understandable that many students do not have the opportunities to publish in top places. Thus, general writings, even those under submissions or rejected, would still be good (and much better than nothing).  But be sure to upload your papers with your application and talk about them in your statement (see \autoref{sec:research-statement}).  Note that local conferences and non-English journals or conferences do
not carry as much weight since their quality is often unknown to US faculty. However, if you have published in such places, you should still mention them in your statement and explain why they are good.

% In CS, conferences are important and top conferences are prestigious. You can find top CS conferences at places such as \href{https://csrankings.org}{CSRankings}, which ranks CS programs based on how their faculty publish at top \emph{conferences} (see more in \autoref{sec:ranking}).

@@

  Sometimes students would go through great length just to get letters from well-known professors in their school (e.g., department head). But as mentioned, if these professors do not know you, the letters are generic and carry little value, in fact, \red{red flags}. Moreover, a top professor in Vietnam might not be well-known to US faculty (see more details in \autoref{sec:your-school}). So save the trouble and just get letters from \emph{any} professors/supervisors who knows you well and can write a good letter about \emph{your} research ability. It's better to have a good personalized
  letter about your own research ability from someone who is less
  well-known than a generic/weak letter from a well-known person.

@@

While you might not have control over LORs or \hyperref[sec:your-school]{where your go to school}, you do over your
statement of purpose (SOP) or personal statement\footnote{Some schools separate these documents and ask you to write both: SOP, which focuses on research experiences, and Personal Statement, which is everything more personal, e.g., why PhD, challenges, diversity, etc}! So write it well because it could make a big difference.
In your statement, you have the opportunity to make your application stand out and unique, which can make you \emph{fit} the CS program you're applying to, even if you don't have very strong research experience.
A well-written SOP also shows that you can communicate, which is very important in research, and that you can effectively teach and communicate with students, which is important for TA funding (see \autoref{sec:funding}).
% is important because if you need (GTA) funding, it will provide evidence
% that you can teach and communicate with students.

There are various guides on writing statement, e.g.,~\cite{blattman2022writing}, and \href{https://cs-sop.org/}{many example statements} are available. So I will not talk too much about statements. In short, focus on your research goal and vision, and convince us that you can achieve it through your experience, e.g., published papers, or if you work on some projects by yourself, talk about it. Also, talk about things that adcom members might not know about, e.g., your Github project with 1K+ stars or your regular contributions to well-known open-source projects. Finally, talk about what makes you different than the other 1000 applicants that we are considering (e.g., see \autoref{sec:improve-your-chance} for increasing your admission chance).

@@

\begin{commentbox}[Vu:]
  I often read the research statement (and the LORs) first. If I am
  persuaded by then, I would skim over other factors and advocate for
  admission (unless I see red flags in other parts). If I am not
  convinced, then I will likely recommend rejection (unless I see
  something standout in other parts).
  \tcblower
  Do careful research on professors, don't mention \emph{emeritus} or  adjunct faculty (see more about various types of faculty in \autoref{sec:faculty-types}).
  Also, be careful not to send statements to wrong schools or mixing
  facts (e.g., talking about school X but mentioned about working with
  profs. at school Y; and definitely do not talk about George Washington when applying to George Mason). I have seen such statements more time that I
  should.
\end{commentbox}

@@

Graduating from top universities \emph{that we recognize} helps. For example, if your school is well-known, then it is \emph{``top foreign''}, which is definitely a plus.
However, if we do not know much about schools in your country, then we are uncertain about the quality of your school and  likely treat your school as \emph{``unknown foreign''}, which can be a minus point.

@@

This can be an issue for students in many top international universities where the competition is so high that very good students can still have low rankings from these schools (and be overlooked by the adcom).
So what to do with this? as mentioned in \autoref{sec:your-school} you can put a note about this in your statement and ask your LoR writers to talk about it.

@@

In short, getting an interview is a good sign; it means that someone is considering you. If we are not interested in your application, we will not proceed with an interview.
However, do not worry much if you do not get an interview! For example, at GMU, most of our admitted students with TA do not go through any interviews.

@@

        You can then share this sheet with your reference writers and remind them to submit LoRs on your behalf (see \autoref{sec:lor}).
\end{itemize}

@@

By around March you should hear back from most PhD programs you applied to. If you haven't heard back, reach out via email and ask about the status of your application.
If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!  Important factors to consider are the reputation of schools and professors (\autoref{sec:schoolsandprofs}), and funding availability (\autoref{sec:funding}). You will have to make your decision by around \emph{April 15}.

@@

As mentioned, if you're admitted to a \emph{good} CS PhD program, you should not have to worry about funding!  
In the US, the common types of funding for PhD are \emph{graduate teaching assistant} (GTA or TA), \emph{graduate research assistant} (GRA or RA), and \emph{Fellowship}.
RA is paid by a prof. for you to do their research. TA is paid by the department for you to help with teaching. Finally, fellowship is an independent funding that can come from a school, a company, or an organization. \autoref{tab:funding} summarizes the differences.
Note that funding is typically more available for PhD students than 
Masters. %Undergraduates typically have no funding and also have to pay international tuition, which is very expensive!.

@@

For a full breakdown of how much a graduate student costs, see \autoref{sec:ra-cost}.

@@

\subsection{Research Assistant (RA)}
RA is provided through a professor through their own funding so you can work on their project.  
You do not need to teach as an RA, so you can focus on your research. Depending on the professor, RA may be available during the summer. \autoref{sec:ra-cost} gives more details on RA budget.

\textbf{How to get RA?} When a professor recruits you, they will likely give you RA right away (e.g., when you apply).  A common scenario is that you first get admitted with TA, and then after a year or two find an adviser to support you with RA. See how to contact a prof. in \autoref{sec:contact} for research opportunities.

@@

Thus, if your goal is Ph.D. in CS, you have to aim only for schools offering such a degree.  %This might sound obvious but many students (even those in the US) get confused due to the large number of universities. 
While this can be confusing due to the large number of universities in the US, a little research, e.g., searching for PhD in CS from the school website, will help you find out. All schools listed in \autoref{sec:ranking} have PhD studies in CS, so you can start there.

@@

You can learn about CS programs and research expertise of faculty using resources such as \href{https://csrankings.org}{CSRankings.org}, which is designed specifically to help prospective PhD students in Computer Science!  You will be very surprised to learn that a school that you didn't know much about can have very strong research in your interested topic (and vice versa, a school you thought highly about might have no faculty working in the research field you're interested in). This is also a good way to learn about individual faculty (who works on what) and well-known CS conferences\footnote{In CS (and probably only in CS), conferences, not journals, are often the main venue to publish research finding.}. \autoref{sec:ranking} gives the top 50 CS programs in the US according to CSRankings.

@@

However, in general, rankings can be superficial and you need to do more research to be informed and make better decision. For example, if you get admissions to several places, you should consider attending Open Houses and contact profs. that you're interested in at those place and talk to them. They would be more willing to chat with you now that you have been admitted.  Ask them questions about their work, how they manage students, their expectations. You can even ask to contact their students. See more in \autoref{sec:accepted} on what to do after getting admitted.

@@

\section{Choosing an Adviser}
Obviously, there is no one-size-fit-all answer to this question. The best adviser is the one that you can work well with, i.e., \emph{fits} you.  This is difficult and takes time. Fortunately, unlike many countries that require finding an adviser and research topic before starting the PhD (\autoref{sec:non-us-differences}), CS PhD programs in the US will typically give you a couple of years to search for advisers and research topics.  This is especially true if you're admitted with TA (\autoref{sec:ta}), which gives you time to explore and find adviser.

@@

Here are some general advice to find professor.  First, search for profs. that share similar research interests. You can learn about faculty and their level of research activity through the faculty's website or even from CSRankings.  For example, in CSRankings, if you want to work with PL, you can search for those publish in PL conferences.  If you want to work with SE \emph{and} AI, you can search for faculty who work in both SE and AI.   You can also go to their website, look at their papers and projects.  In any case, you should contact the professor and talk to them. For example, you can (cold) email them and ask about their research and if they are taking new students (\autoref{sec:contact}), and schedule appointments to chat with them (office hrs are often the best time).

@@

  In other fields, e.g. Software Engineering, Vietnamese students face challenges in reaching US professors. Do you have any tips for Vietnamese students who want to connect with US professors and work as research assistants?
  \tcblower
  \textbf{Vu}: \autoref{sec:contact} shows how to contact a professor for research opportunities. Many probably will say no (or do not reply) as they do not have the bandwidth to take on random students, but some may say yes if they see potential fit.
\end{commentbox}

@@

Not every faculty can serve as your formal adviser. Let's try to understand the different types of faculty and their roles. For example, you might hear about tenured,  tenure track, and teaching faculty.  You might also hear about assistant, associate, full, adjunct, emeritus, teaching, research professors. Here is a primer on these terminologies.

\paragraph{Tenure-line and teaching faculty} Tenure line faculty, consisting of tenured and tenured-track profs., focus on research, which includes publishing papers, obtaining grants, mentoring Ph.D students.  They often have very low teaching load (e.g., 1 per semester). \emph{Tenured} faculty are professors who have been promoted to have a permanent position (informally, very hard to fire them).   \emph{Tenure-track} faculty are (more likely young) professors who are on the track to get tenure.  \textbf{Both tenure and tenured track faculty can serve as a formal adviser of Ph.D. students}. This is very important because you need to have a tenure line faculty as your main PhD adviser. \autoref{sec:tenure-vs-tenure-track} talks more about choosing between tenure and tenured-track professor as your adviser.

In contrast, \emph{teaching} (or instructional faculty or professor of practice) mainly focus on teaching. They typically teach 3--4 classes per semester (which is quite a lot) and \emph{do not} have research responsibilities (e.g., they do not have to worry much about publishing papers or obtaining grants).  They can mentor Ph.D. students but they typically cannot serve as a formal adviser of Ph.D. students (i.e., they \emph{cannot graduate} PhD students).

It is worth noting that professors in a non-CS department also unlikely can serve as formal PhD advisers for CS students (see \autoref{sec:related-fields}).

\paragraph{Titles} Faculty have rankings, e.g., assistant, associate, full, regardless if they are tenure line or teaching.  Assistant means new faculty, associate means they have been promoted, and full means they are senior. A tenure track faculty starts with being an assistant professor and then gets tenure and promoted to associate (typically after 6 years) and then full (time varies, some become full within 3-4 years, some 10+ years, some remain associate).

Adjunct faculty is not full-time, e.g., they might be working in industry and teach a class or two for fun.  Emeritus means they are retired but still have some affiliation with the university.  Research faculty (or research scientist) are typically non-tenure line faculty who focus on research.  Due to their roles, adjunct, emeritus, and research faculty typically do not advise Ph.D. students.

@@

First, if you want to contact a prof. to \emph{ask about your admission chance}, please \textbf{don't}. We don't know and can't answer because as explained in \autoref{sec:evalapps}, we don't make individual decisions and might not even be assigned to evaluate your application.  It is the same as sending a paper draft to a journal editor and ask them if your paper has a chance.
